---
title: Writing challenge day 28
date: 2018-09-27
categories: ["research", "writing", "learning", "DocumentingLearning"]

---

This month has been fast. The writing has been really interesting, I was able to get some old stories written and had a lot of fun doing it. I would like to continue with the habit, it is already making it easier for me to put words into text. I'm not sure about the quality, but I experience the *writing ease* for both Spanish and English.

## The data

```{r}
#| label: setup
#| echo: false
#| results: hold
#| message: false
#| warning: false
library(ggplot2)
library(dplyr)
suppressPackageStartupMessages(library(lubridate))
```

I had some pretty long bouts on Saturday and Sunday, writing close to 3000 words in aggregate. Of course, considering my overall trend, these instances look like outliers, but it didn't feel that way. 

It felt like a reasonable production, given the fact I was: 1) not tired, 2) somewhat isolated from the outside world, and 3) in a writing mood.

```{r}
#| echo: false
#| message: false
#| warning: false

tracking_writing <- read.csv("tracking_writing.csv")

tracking_writing <- dplyr::select(tracking_writing, -place) %>%
  mutate(date=ymd(date)) %>% 
  filter(date < as.Date("2018-09-27"))

ggplot(tracking_writing, aes(date,word_count))+
  geom_point()+
  scale_x_date(date_labels="%b %d",date_breaks  ="5 day")+
  geom_smooth(color="black")+
  theme_bw()+
  labs(title = "Word count per writing bout", x = element_blank(), 
       y = element_blank())

```

The aggregate trend has recovered a bit from a sloppy third week of the challenge. The cumulative average is `r tracking_writing %>% group_by(date) %>% summarise(total_count = sum(word_count)) %>% summarise(mean_count = round(mean(total_count,na.rm=TRUE),2)) %>% pull()`, which sounds way better than *just* achieving 50% of the target. Still, ~1000 words a day is a solid number if I'm able to consistently put that amount of words outside my head and into a product. 

```{r}
#| echo: false
#| warning: false
#| message: false

tracking_writing %>% group_by(date) %>%
  summarise(total_count = sum(word_count)) %>%
ggplot(aes(date,total_count))+
  geom_hline(yintercept = 2000, lty=2)+
  scale_x_date(date_labels="%b %d",date_breaks  ="5 day")+
  geom_smooth(color="black")+
  geom_point()+
  theme_bw()+
  annotate(geom="text", label="daily target", x= ymd("2018-09-11"), y=2100)+
  labs(title = "Total Word Count per Day", x = element_blank(),  y = element_blank())



```

## General comments

It's been a great month of writing. I took a challenge knowing that it was going to be tough and interesting. It has certainly been a great way to force me to put words into text.

I intend to continue pushing myself to write. I definitely should say that having a high bar is a great driving force, and knowing that I'm really unlikely to get 2000 words unless I sit to write several hours undisturbed is somewhat comforting. I would say the pressure to produce is at a good level and I'm confident I can maintain/slightly increase this amount of production. 

Solid work on the outline is hands down the best first step. Initial ideas or sub-headers can grow into a paragraph and, a coherent story starts to materialize before your eyes. The outline still needs good content to back it up, having a bunch of content/ideas is the key to write a good amount of quality text. 

## The good

I finished another short story, two blog posts and was pretty close of hitting the word count on two days. I was afraid about the downward trend but I reversed bad third week and transformed it into a (so far) better fourth week. Weekend writing is the best!

After a month, Sublime Text 3 and I are best friends. I use it for `.csv`, `.md` and sometimes even `.Rmd` files. I go back and forth between Sublime and Rstudio or couple Sublime with pandoc via command line. I found this writing experience to be better than what I had with Word. I specially like the aesthetics of the program, the dark background doesn't mess with my eyes[^Rstudio] and I really enjoy the different functions to play around.

I know this won't happen but I am really over with Word and `WYSIWYG` programs (yes Excel, I'm talking to you too, you started all this!). The truth is that most people will continue to use them or find them more convenient. Hence, until a critical mass outside these type of programs is not reached, everyone will keep using them to write and share documents. Hopefully, with time and the new generations being taught how to program in school, Markdown (or something like it) will be more accessible and we will overcome this problem[^Iwonder].

## To improve

Working *before* writing the main portion of text is key. Having the ideas out is important but they don't have to be fully materialized. Doing the work of outlining where the text will go can get you a long way. It can give you a hard substrate to anchor your writing bouts from. Connecting the dots is a separate work, you can't flow if you are busy editing a sentence while you write it. First, flow, get the ideas out, *all of them*. Later connect and edit.


[^Rstudio]: Although for some reason I can't get myself to move away from the white (default) version of Rstudio.
[^Iwonder]: I wonder if I'm going to be 80 by then and not willing to learn new things.
