---
title: "XY-density-maps"
date: 2018-07-13
categories: ["R"]
execute: 
  echo: true
---

I have been playing around with tracking software and the ways to visualize the position of an animal in an arena over time. Even with normal cameras (30 Hz) and relatively small periods of time (~5 min) we get massive datasets and the only way to make sense of them is to aggregate data over time. I have been interested in neat visualizations (using R), thus here are some ways/comments. I will explore different packages and doing comparisons of the results.


## Packages

```{r}
#| label: load
#| warning: false
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(KernSmooth))
suppressPackageStartupMessages(library(raster))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))

```


## Data

I have an example dataset with `XY` position of the animal. I also have other morphological data that is irrelevant for this case. See example below: 

```{r}
#| label: dataset
#| echo: false


rat_pos <- read.csv('2018-03-20-XY-density-maps-files/rloess_smooth.csv')

head(rat_pos)

```


I will do some setup that is common for all the graphs (aka, color palette and the dimensions of the arena).

```{r}
#| label: setup-bins
# Get color palette 

refCol <- colorRampPalette(rev(brewer.pal(6,'Spectral')))
mycol <- refCol(6)

# define bin sizes
bin_size <- 40

# Camera resolution is 640x480. Hence...
xbins <- 640/bin_size 
ybins <- 480/bin_size
```

## ggplot2

The reason I always go to `ggplot2` first is because it's awesome, I buy into the grammar and find it intuitive to accumulate layers over layers. The underlying thought is that `ggplot2` handles all problems. In this case, the result has some pros (layers, layers and more layer), and some cons (basically, it doesn't look amazing). 

```{r}
#| label: ggplot2-option-1
p <- ggplot(rat_pos, aes(X,Y)) + 
  stat_density2d(geom = 'tile', 
                 aes(fill = after_stat(density)), 
                 contour = FALSE,
                 n = c(xbins, ybins)) +
  coord_equal() +
  theme_minimal() +
  scale_fill_gradientn(colors =  mycol) +
  geom_vline(xintercept = c(0,640))+
  geom_hline(yintercept = c(0,480))

print(p)
```

This is interesting because we can overlay things into the plot. For example the trace:

```{r}
#| label: ggplot2-option-1-trace

p + geom_path(alpha=0.1)

```


We can further remove the axis (or any other modifications we feel like doing). 

```{r}
#| label: ggplot2-option-1-noborder

p + geom_path(alpha=0.5) + theme_void()

```

### Helping ggplot from outside

I found that, if we calculate the density externally, it looks smoother. This is a mixed, `bkde2D` mediated, `ggplot2` approach (aka the best of 2 worlds).

```{r}
#| label: ggplot2-option-2
bins <- bkde2D(as.matrix(rat_pos[,1:2]), 
               bandwidth = c(xbins, ybins),
               gridsize = c(640L, 480L))

# reshape
bins_to_plot <- tidyr::pivot_longer(
  data.frame(bins$fhat) %>% tibble::rownames_to_column(), 
  cols = -rowname) %>% 
  # fix the name column
  mutate(name = stringr::str_remove(name, "X")) %>% 
  # convert to numeric
  mutate_all(as.numeric)


ggplot(bins_to_plot, aes(rowname, name, fill = value)) +
  geom_raster()+
  coord_equal() +
  theme_minimal() +
  scale_fill_gradientn(colors =  mycol) +
  geom_vline(xintercept = c(0,640))+
  geom_hline(yintercept = c(0,480))+
  theme_void()


```

### 2023 Update

As of 2023, ggplot is able to handle 2D density heatmaps much better. I haven't played too much with it, but here are some examples

```{r}
ggplot(rat_pos, aes(X,Y)) + 
           stat_density_2d(geom = "polygon", 
                           contour = TRUE,
                           aes(fill = after_stat(level)),
                  bins = 8)+
  coord_equal() +
  scale_fill_gradientn(colors =  mycol) +
  theme_minimal() +
  geom_vline(xintercept = c(0,640))+
  geom_hline(yintercept = c(0,480))+
  theme_void()

```

```{r}
ggplot(rat_pos, aes(X,Y)) + 
  geom_density_2d_filled(bins = 8)+
  coord_equal() +
  theme_minimal() +
  scale_fill_brewer(palette = "Spectral", direction = -1)+
  geom_vline(xintercept = c(0,640))+
  geom_hline(yintercept = c(0,480))+
  theme_void()

```


## Using raster package

The `raster` package is maybe an older solution, which is surprisingly low demand. In 3 lines of code we get a perfectly functional plot. On the other hand, it's not the best looking graph and we get the caveats (yes, `base` graphics).   

```{r}
#| label: raster-package
r <- raster(xmn=0, ymn=0, xmx=640, ymx=480, res=20)
x <- rasterize(rat_pos[,1:2], r, fun='count')
plot(x, xlim=c(0,640), ylim=c(0,480), xaxt='n', ann=FALSE, yaxt='n')
```


## Documentation 

I have found a good amount of advice and inspiration in the links below. 

* http://stat405.had.co.nz/ggmap.pdf
* https://stackoverflow.com/questions/24078774/overlay-two-ggplot2-stat-density2d-plots-with-alpha-channels?lq=1
* https://www.r-bloggers.com/5-ways-to-do-2d-histograms-in-r/
* https://stackoverflow.com/questions/24845652/specifying-the-scale-for-the-density-in-ggplot2s-stat-density2d
