---
title: "Writing Checklist"
date: 2020-01-02
categories: ["writing", "learning", "DocumentingLearning", "R"]
execute: 
  echo: true
---

I spent a good portion of the end of 2019 and the beginning of 2020 reading Paul Graham's essays. Many caught my attention (and my fingers to take notes). I want to comment on an essay about advice for good writing. Because it was written in a way that allowed for easy reformatting (ideas separated by `;`), I thought it would be cool to convert the prose into a checklist).  

## Creating the checklist

The main text comes from [here](http://www.paulgraham.com/writing44.html). For simplicity, I have only used the main paragraph (aka, copy and paste the second paragraph). I omitted it here because it would be long and you can read it in the native format from the source.

```{r}
#| echo: false
text <- c("Write a bad version 1 as fast as you can; rewrite it over and over; cut out everything unnecessary; write in a conversational tone; develop a nose for bad writing, so you can see and fix it in yours; imitate writers you like; if you can't get started, tell someone what you plan to write about, then write down what you said; expect 80% of the ideas in an essay to happen after you start writing it, and 50% of those you start with to be wrong; be confident enough to cut; have friends you trust read your stuff and tell you which bits are confusing or drag; don't (always) make detailed outlines; mull ideas over for a few days before writing; carry a small notebook or scrap paper with you; start writing when you think of the first sentence; if a deadline forces you to start before that, just say the most important sentence first; write about stuff you like; don't try to sound impressive; don't hesitate to change the topic on the fly; use footnotes to contain digressions; use anaphora to knit sentences together; read your essays out loud to see (a) where you stumble over awkward phrases and (b) which bits are boring (the paragraphs you dread reading); try to tell the reader something new and useful; work in fairly big quanta of time; when you restart, begin by rereading what you have so far; when you finish, leave yourself something easy to start with; accumulate notes for topics you plan to cover at the bottom of the file; don't feel obliged to cover any of them; write for a reader who won't read the essay as carefully as you do, just as pop songs are designed to sound ok on crappy car radios; if you say anything mistaken, fix it immediately; ask friends which sentence you'll regret most; go back and tone down harsh remarks; publish stuff online, because an audience makes you write more, and thus generate more ideas; print out drafts instead of just looking at them on the screen; use simple, germanic words; learn to distinguish surprises from digressions; learn to recognize the approach of an ending, and when one appears, grab it.")

```

In the code chunk below I comment the steps I took.

```{r}
# text is copy-paste of the second paragraph
# text <- c("...") ... == copy-paste

# split using "; "
sp <- stringr::str_split(text, "; ")

# To create a checklist
# DASH SPACE [ ] SPACE SPACE
sp <- paste0("- [ ]  ", unlist(sp))
# append markdown title
sp <- c("# Paul Graham's Writing Checklist", sp)

# Capitalize first letter of every sentence
sp <- stringr::str_to_sentence(sp)
```

If you want to write it to file, you can use `writeLines()`

```{r}
#| eval: false
# write
fileConn <- file("paul_graham_writing_checklist.md")
writeLines(sp, fileConn)
close(fileConn)

# If you want to knit to html you can do this
# knitr::knit2html("paul_graham_writing_checklist.md")

```

Let's take a look at the product:

```{r}
# printing for web post
cat(sp, sep = "\n")
```

The checklist is not in strict chronological order, so I might try to reshape it later into something that would make more sense as a timeline. It will render as a checklist in something that is GitHub flavored (e.g., on GitHub). However, I will probably just use it as text file or print it as is. 

I also noticed that the link for the Spanish translation is broken/outdated, so I tried my best to translate it. [Here it is](/2020/01/escribiendo-en-pocas-palabras/)

## Summary

I think this checklist is an awesome learning opportunity and a clear path to improving one's writing. Quite happy to have found it, looking forward to improving my skills.

I got to learn a few cool R things like, 1) I can use `writeLines()` to write a `.md` file and 2) `stringr::str_to_sentence()` is an **awesome** function.
