---
title: "Original Text"
description: "Checking how the use of verbatim quotes in books"
date: "2023-11-07"
categories: ["R",]
code-fold: true
echo: true
---

```{r, include = F}
library(tidyverse, warn.conflicts = F)
library(ggethos)
```

I have been lately noticing a glaring trend in some of the non-fiction books that I read: the use and abuse of verbatim quotes. They come in the shape of:

> "As such and such said: INSERT LONG VERBATIM TEXT HERE"

Of course, there are no rules regarding the use of verbatim text[^copy]. But, if I can get a sense of overuse only from reading the book, it makes me curious to go get look at the data.

How much of the book is actually a verbatim text dump? Would I bet is 10%? Maybe 20%? Would lower percentages make me go easier on the author or is this a lost cause (i.e., if I notice the overuse by reading, all hope is lost)?

## Example Book

Enough of chatter. Let's try to answer this by analyzing one of the books in question: "Do Nothing" by Celeste Headlee. Reading the book in R using the `epubr` package gives us this table:

```{r}
# read the epub
book_text <- epubr::epub("Do Nothing - Celeste Headlee.epub")
book_text$data[[1]] %>% 
  mutate(text = str_sub(text, 0, 20),
         text = paste(text, "...")) %>% 
  gt::gt() # needs to get the text stings truncated
```

We can get rid of the legal stuff that normally goes before the text and everything that comes after the content (i.e., acknowledgements and references).

```{r}
# A simple slice operation would do
book_txt <- book_text$data[[1]] %>% 
  slice(6:24)
```

We can also get some metadata from the text (will come useful for later).\

```{r}
# bind previous word and character counts
meta <- book_txt %>% 
  select(section, nword, nchar) %>% 
  mutate(part = paste0("part", 5:23)) %>% 
  select(-section)

```

Now that we have the text, we can find all the instances of `"something in between these quotes here"` using `stringr::str_locate_all():`

[^copy]: But I'm sure a copyright lawyer would know much more than I do regarding how much verbatim text you can include and still claim ownership of your work.

```{r}
# extract the text
match_df <- stringr::str_locate_all(book_txt$text, '"(.*?)"') %>% 
  # give names for future binding
  # parts go from 5 to 23 (idx goes 6:24)
  set_names(nm = paste0("part", 5:23)) %>% 
  # convert into tibble for easy binding
  map(as_tibble) %>% 
  bind_rows(.id =  "part")

```

Below, I'm showing a slice with an example of matched character positions and how they would look like in the text. I want to direct your attention to the second and third row. I hope you notice that these two quotes are, in fact, one single quote that was split into two.

```{r show-matches}
# This is an example
match_df %>% 
  slice(8:10) %>% 
  mutate(quote = map2_chr(
    start, end,  
    function(.x, .y) str_sub(book_txt$text[[1]], .x, .y)
  )) %>% 
  gt::gt() %>%
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
        )
```

### Merging Quotes

The issue of quotes being split arises not because of a bug in code, but because the author writes in this way. She would do something like:

> "A palm tree", somebody said, "belongs to the Plant Kingdom."

These stylistic choices will modify the statistics for the direct quotes (e.g., the average length of a quote will be much lesser than if these quotes were kept verbatim). I decided that I want to merge quotes if they are too close to each other (I will try 100 characters[^arbitrary]). This will slightly inflate my % counts, since I'm attributing characters that are not direct quotes to actual quotes. Thus, when I calculate percentages, I will do so without merging (see @percentages-with-no-merging).

[^arbitrary]: Of course, this threshold is arbitrary. How did I come up with it? I asked ChatGPT to come up with 10 interjections that were a bit longer than "they said" and phrases where sitting comfortably around 50. I doubled it to be super sure that we were not missing instances.

There's one neat trick using `lag` and `cumsum` with a condition to achieve conditional grouping. We can see that rows 9 and 10 are marked as belonging to the same group now 🎉.

```{r}
threshold <- 100  # Define your threshold

merged_quotes <- match_df %>%
  mutate(
    .by = part, 
    prev_end = lag(end),
    distance = start - prev_end,
    merge_group = cumsum(ifelse(is.na(distance) | distance > threshold, 1, 0))
  ) 

# 
head(merged_quotes, n = 10)
```

This intermediate step also gives us the answer to a new question:

> What is the average distance between quotes?

```{r, include=FALSE}
m <- mean(merged_quotes$distance, na.rm = T) %>% signif(2)
sdd <-  sd(merged_quotes$distance, na.rm =T) %>% signif(2)
m_sd <- glue::glue("x̄= {m} ± sd = {sdd}")
# average chars per word
chars_per_word <- mean(meta$nchar / meta$nword)
```

The answer is `r m_sd` . On average, you start a new qoute after `r signif(m / chars_per_word, 2)` words of original content. Is that a lot? Is that too little?

To be honest, it feels true to the reading experience. My sensation was that the author was using the verbatim quotes with high frequency, and the data seems to align with that. But don't take my word for it, let's try to visualize it.

We are two steps away from the viz.

1.  Do the actual merge
2.  Add the end of each chapter

We can do `Step 1` using the code below:

```{r}
merged_quotes <- merged_quotes %>%
  summarize(
    .by = c(merge_group, part),
    part = first(part),
    start = first(start),
    end = last(end)
  ) %>% 
  # add the lag again to see where the original text starts
  mutate(text_start = lag(end, default = 0), .by = part)
  
```

Right now, we have the start of the original text in `text_start` and the `start` and `end` of each verbatim quote. We need to make use of the metadata stored in `meta` to add the end of the original content for of each chapter. This only matters for the very last portion that we are going to plot, so I will make a new data set that contains those values instead of merging everything together. To visualize it, I'm going to make use of a package I developed called `ggethos`. You can check it out [here](https://matiasandina.github.io/ggethos/) or adapt the code to work with `geom_segment()`.

```{r}
# pad parts  for plotting
format_part <- function(part_name) {
  # Extract the numeric part
  part_number <- as.integer(str_extract(part_name, "\\d+"))

  # Pad the number with zeros and prepend 'part'
  formatted_part <- str_c("part", str_pad(part_number, width = 2, pad = "0"))
  
  return(formatted_part)
}

# make tail end segments
tail_data <- merged_quotes %>% 
  summarise(.by = part, last_quote_end = max(end)) %>% 
  left_join(meta, by='part') %>% 
  # fix the padding after merging
  mutate(part = format_part(part))

# fix the padding here too
merged_quotes <- merged_quotes  %>% mutate(part = format_part(part))


ggplot(data=merged_quotes) + 
  geom_ethogram(aes(x=text_start, xend=start, y = part), color ="gray30") +
  geom_ethogram(data=tail_data, aes(x=last_quote_end, 
                                    xend=nchar, y = part), color ="gray30") +
  geom_ethogram(aes(x=start, xend=end, y = part), color = "red")+
  cowplot::theme_nothing() +
  labs(title = "'Do Nothing' is Peppered by Quotes",
       subtitle = "<span style = 'color:gray30'>Original Text</span> and <span style = 'color:red'>Verbatim quotes</span>",
       caption = "Viz: Matias Andina",
       y = "Chapter") +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = ggtext::element_markdown(hjust = 0.5),
    plot.background = element_rect(fill = "black"),
    text = element_text(color = 'gray80'),
    axis.title.y = element_text(angle = 90),
    plot.caption = element_text(size = 8, hjust = .95))
```

I believe this plot conveys a good mental image of what reading the book feels like in terms of verbatim text usage.

### Percentages with no merges

As mentioned in the beginning of the article, I was curious about *how much* verbatim text there was. Again, using the number of characters in each chapter stored in the `meta` object, we can easily calculate the percentage of all characters that are directly quoted:

```{r}
match_df %>% 
  mutate(quote_chars = end - start) %>% 
  summarise(.by = part, 
            quote_chars = sum(quote_chars)) %>% 
  left_join(meta, by = "part") %>% 
  mutate(quote_frac = quote_chars / nchar,
         part = fct_reorder(part, quote_frac)) %>% 
  ggplot() +
  geom_hline(aes(yintercept = mean(quote_frac)), lty = 4) +
  geom_point(aes(as.numeric(part), quote_frac), 
             size = 4, alpha = 0.9, color = "darkred") +
  geom_label(aes(x = 15.5, y = 0.17,
                 label = paste(part[which.max(quote_frac)],
                               scales::percent(max(quote_frac)),
                               sep="\n"
                 ))) +
  scale_y_continuous(labels = scales::label_percent(),
                     expand = expansion(add = c(0.01, 0.05)))+
  labs(y = "Verbatim Quotes",
       x = "Book Part\n(ascending quote % order)",
       title = "'Do Nothing' contains ~10% verbatim quoted text",
       subtitle = "Some parts are as high as 17%!")+
  cowplot::theme_minimal_hgrid()
  
```

## A silver lining

Most non-fiction books are a regurgitation of something somebody else said a long time ago (there's nothing new under the sun). In a sense then, it's more truthful for an author to quote verbatim from the original source than to paraphrase whatever they took out of it and hide the initial message under a footnote[^meta].

[^meta]: This paragraph was indeed a paraphrase of my editor's (read wife's) reaction to my article. Talking to her is a great exercise in positive reframing.
